---
title: Creating Presence Absence Matrices for Costa Rican Birds
author: "Motmots Lab Group"
date: "18 September 2021"
output: rmarkdown::github_document
---

# 1 Overview

This study uses ecological niche models created using minimum volume ellipsoids by MF Velde. These models are spatially restricted around two test localities for a different concurrent project; these restrictions will have little to any effect on the studies performed herein.

First, we need to load the relevant packages.

```{r,eval=F}
library(ape)
library(data.table)
library(dismo)
library(ggplot2)
library(raster)
library(rgdal)
library(rgeos)
library(sf)
library(tidyverse)
library(vegan)
library(velox)

# also cite ggplot2, loaded with ecostructure
```

```{r,echo=F}
library(ape)
library(data.table)
library(dismo)
library(ggplot2)
library(raster)
library(rgdal)
library(rgeos)
library(sf)
library(tidyverse)
library(vegan)
#library(velox)

# also cite ggplot2, loaded with ecostructure
```

# 2 Winter Models

## 2.1 Creating training areas

The MVE models were created, trained, and projected to the entirety of Costa Rica and Panama. For this study, we are explicitly looking at Costa Rica, given the density of data and how well known the bird communities are in the country. Additionally, the MVE models were projected out without regard to biogeographic barriers. Here, we are correcting for these changes and ensuring that all species are restricted to their appropriate biogeographic zones to create more accurate species distribution models from which other metrics can be calculated.

*Note*: we used broad biogeographic zones as fine scale biogeographic partitioning of the pacific slop heavily biased models. So we are using **five training areas**, as follows:

1. Atlantic slope
2. Pacific slope (which includes all but the following:)
3. Nicoya Peninsula
4. Osa Peninsula and adjacent coast

We create different areas for the peninsulas as these highlands are not connected to the main cordilleras. We refrained from splitting the northern and southern mountains as many species seem to be able to cross this gap, and even supposed endemic (*e.g., Elvira cupreiceps*) appear to make it across the central valley break.

We will later be converting these data to 10 km grids to facilitate the usage of ECOSTRUCTURE in the landscape.

```{r,echo=F}
sdm.filepath=paste0("~/Dropbox/motmots/SDMs/MadreSelva_december/")
shp.filepath=paste0("~/Dropbox/motmots/shapefiles/")
```
```{r}
# get list of files with points

gpkgs=list.files(shp.filepath,pattern=".gpkg")

y=readOGR(paste0(shp.filepath,gpkgs[2]))

crs.x=crs(y)

plot(y)
```

An example of one of the training areas; this is the Osa Peninsula.

```{r}
# We create training areas based on winter data for most
# winter is the time of dispersal and vagrancy for many taxa
# summer species not dealt with here are dealt with later.

species=list.files(paste0(sdm.filepath,
                          "all.species"),
                   pattern="*.csv")
```
```{r,eval=F}
# preload shapefiles since we have only four

y1=readOGR(paste0(shp.filepath,gpkgs[1]))
y2=readOGR(paste0(shp.filepath,gpkgs[2]))
#y3=readOGR(paste0(shp.filepath,gpkgs[3]))
#y4=readOGR(paste0(shp.filepath,gpkgs[4]))

for(i in 1:length(species)){
  x=read.csv(paste0(sdm.filepath,"all.species/",species[i]))
  x2=x%>%dplyr::select(LONGITUDE,LATITUDE)
  x.data2=SpatialPoints(x2,proj4string = crs.x)
  
  shp.list=NULL
  
  # for variable number of gpkgs files
  
  #for(k in 1:length(gpkgs)){
  #  y=readOGR(paste0(shp.filepath,gpkgs[k]))
  #  inside.m=sum(!is.na(over(x.data2,as(y,"SpatialPolygons"))))
  #  if(inside.m>0){shp.list[k]=1}else{shp.list[k]=0}
  #}
  
  # for this instance, since there are only five
  
  shp.list[1]=sum(!is.na(over(x.data2,as(y1,"SpatialPolygons"))))
  shp.list[2]=sum(!is.na(over(x.data2,as(y2,"SpatialPolygons"))))
  #shp.list[3]=sum(!is.na(over(x.data2,as(y3,"SpatialPolygons"))))
  #shp.list[4]=sum(!is.na(over(x.data2,as(y4,"SpatialPolygons"))))
  
  shps=which(shp.list>0)
  
  if(length(shps)==0){next}
  
  if(length(shps)==1){
    # x.u.1=readOGR(paste0(shp.filepath,gpkgs[shps]))
    if(shps==1){assign('x.u.1',y1)}
    if(shps==2){assign('x.u.1',y2)}
    #if(shps==3){assign('x.u.1',y3)}
    #if(shps==4){assign('x.u.1',y4)}
  }else{
    # merge shapefiles
    #x.u.1=readOGR(paste0(shp.filepath,gpkgs[shps[1]]))
    #for(k in 2:length(shps)){
    #  x.u.2=readOGR(paste0(shp.filepath,gpkgs[shps[k]]))
    #  x.u.1=gUnion(x.u.1,x.u.2)
    #}
    
    # memory saver below
    for(k in 1:length(shps)){
      if(k==1){
        if(shps[k]==1){assign('x.u.1',y1)}
        if(shps[k]==2){assign('x.u.1',y2)}
        #if(shps[k]==3){assign('x.u.1',y3)}
        #if(shps[k]==4){assign('x.u.1',y4)}
        #if(shps[k]==5){assign('x.u.1',y5)}
      }else{
        if(shps[k]==1){assign('x.u.2',y1)}
        if(shps[k]==2){assign('x.u.2',y2)}
        #if(shps[k]==3){assign('x.u.2',y3)}
        #if(shps[k]==4){assign('x.u.2',y4)}
        #if(shps[k]==5){assign('x.u.2',y5)}
        
        x.u.1=gUnion(x.u.1,x.u.2)
      }
    }
  }
  
  nombre=strsplit(species[i],"[.]")[[1]][1]
  
  x.union=st_as_sf(x.u.1)
  
  st_write(x.union,
           paste0(shp.filepath,"training_areas/",nombre,".gpkg"),
           nombre)
}

print('done')
```
## 2.2 Clipping Rasters

Next, we need to clip all rasters to the aforementioned training areas and ensure that we have the proper biogeographic envelopes applied for each species.

```{r,echo=F}
shp.path="~/Dropbox/motmots/shapefiles/training_areas/"
sdm.path="~/Dropbox/motmots/SDMs/MadreSelva_december/"
```

Note the SDM subdirectories are `output_all/75` for the whole files thresholded with 75% confidence and `clipped_SDM` for the clipped rasters.

```{r}
sdm.list=list.files(paste0(sdm.path,"output_all/75"),pattern="*.tif")
shp.list=list.files(paste0(shp.path),pattern="*.gpkg")
```

We will not be clipping training areas for each species to correct for species richness patterns etc.

```{r,echo=F}
i=1

name=strsplit(sdm.list[i],"_")[[1]][1]
x=raster(paste0(sdm.path,"output_all/75/",sdm.list[i]))
y=readOGR(paste0(shp.path,shp.list[which(shp.list%like%name)]))

plot(x)
plot(y,add=T)
```

Above is an illustration of `r print(name)`, where the areas where points have been recorded are highlighted in black, and areas outside of the Costa Rica training area or where the species truly does not occur are outside of this biogeographic envelope.

In order to ensure everything is cropped correctly, we will use a 'reference raster' from a widespread species.

```{r,echo=F}
i=which(sdm.list%like%"Amazilia-tzacatl")

name=strsplit(sdm.list[i],"_")[[1]][1]
x=raster(paste0(sdm.path,"output_all/75/",sdm.list[i]))
y=readOGR(paste0(shp.path,shp.list[which(shp.list%like%name)]))

plot(x)
plot(y,add=T)
```

Above is *Amazilia tzacatl*, a widespread Neotropical hummingbird, as predicted by the ecological niche models.Since it occurs in every part of Costa Rica, we can use it to create a default projection raster.

```{r,echo=F}
# scratch space

ex.y=extent(y)
crs(x)=crs(y)
rx=projectExtent(y,crs=crs(x))

x.temp=raster(ext=extent(rx),crs=crs(x),res=res(x))

nu.x=projectRaster(from=x,to=x.temp,method='ngb',
                   res=(res(x)),crs=crs(x))
```

```{r}
# creating template from Amazilia

cr1=crop(x,y)
cr2=raster::mask(cr1,y)
crs(cr2)=crs(x.temp)
cr3=projectRaster(cr2,x.temp,method='ngb',
                  res=(res(x)),crs=crs(x))

cr3[cr3>0]=0

x.temp=cr3

plot(cr3)
```

We can see here that the template extent works. No we can start clipping everything. The template extent importantly has zero values for the entire land, so stacking and summing will create maps that cover the entire country.

Furthermore, we want to create hexbins of everything so we can perform ecostructure analyses.

**Important note**: due to issues with the hex polygons and classes I did all of this in memory at the same time.

```{r,eval=F}
# create hexbin over Zonotrichia training
y2=as(y,"Spatial")
CR.buff=gBuffer(y,width=0.05)
CR.hex=spsample(CR.buff,type='hexagonal',cellsize=0.125)
hex=HexPoints2SpatialPolygons(CR.hex)

Pam.coords=as.data.frame(CR.hex@coords)
Locality=paste0("ID",1:nrow(Pam.coords))

locality_metadata=cbind(Locality,Pam.coords)
colnames(locality_metadata)=c('Locality','Longitude','Latitude')
write.csv(locality_metadata,paste0(sdm.path,"locality_metadata.csv"),
          quote=F,row.names = F)

sf.hex=st_as_sf(CR.hex)
```
```{r,echo=F,eval=F}
write_sf(sf.hex,"~/Dropbox/motmots/SDMs/hex_mesh_slopes.gpkg")
```

```{r,echo=F}
sf.hex=read_sf("~/Dropbox/motmots/SDMs/hex_mesh_slopes.gpkg")
```
```{r,eval=F}
plot(y)
plot(hex,add=T)
```
```{r,echo=F}
plot(y)
plot(sf.hex,add=T)
```

Now we have a hex grid created. Our goals are to create 1) a biogeographic SDM and 2) a hex grid representation of that SDM for the creation of our PAM.

```{r,echo=F}
y=readOGR("~/Dropbox/motmots/SDMs/hex_mesh_slopes.gpkg")
y

ycoord=y@coords

disty=99999999999

for(i in 1:(nrow(ycoord)-1)){
  c1=ycoord[i,]
  for(j in (i+1):nrow(ycoord)){
    disty=c(disty,pointDistance(c1,ycoord[j,],lonlat = T))
  }
}

# convert meters to km for closest distance
disty=disty/1000
disty=disty[-1]
hist(disty)
min(disty)

# calculate area of hexagon
# Area = 1/2 x perimeter x apothem
apothem=min(disty)

# calculate length of 1/12 perimeter from apothem
# since it is 30-60-90, apothem = x*sqrt(3)
halfside=apothem/sqrt(3)
fullside=halfside*2
perim=fullside*6

0.5*perim*apothem
```

```{r,eval=F}
PAM=NULL

errors="Errors"

for(i in 412:length(sdm.list)){
  name=strsplit(sdm.list[i],"_")[[1]][1]

  if(length(which(shp.list%flike%name))<1){
    errors=c(errors,name)
    next
  }
  if(length(which(sdm.list%flike%name))<1){
    errors=c(errors,name)
    next
  }
  
  x=raster(paste0(sdm.path,"output_all/75/",sdm.list[i]))
  y=readOGR(paste0(shp.path,shp.list[which(shp.list%like%name)]))
  
  cr1=crop(x,y)
  cr2=raster::mask(cr1,y)
  crs(cr2)=crs(x.temp)
  cr3=projectRaster(cr2,x.temp,method='ngb',
                    res=(res(x)),crs=crs(x))
  
  # make everything zero so land cells become 0 when stacked
  cr3[is.na(cr3)]=0
  
  x.stack=stack(cr3,x.temp)
  cr4=calc(x.stack,fun=sum)
  
  # performed in previous iteration
  if(file.exists(paste0(sdm.path,'clipped_SDM/',
                              name,"_cropped.tif"))==F){
    writeRaster(cr4,
              filename=paste0(sdm.path,'clipped_SDM/',
                              name,"_cropped.tif"))
  }
                              
  
  # project to hex
  rm(x)
  rm(y)
  rm(cr1)
  rm(cr2)
  rm(cr3)
  rm(cr4)
  
  v.cr=velox(paste0(sdm.path,'clipped_SDM/',
                              name,"_cropped.tif"))
  
  hex.ext=v.cr$extract(hex)
  
  #hex.ext=raster::extract(cr4,hex)
  
  hex.vals=NULL
  
  for(k in 1:length(hex.ext)){
    n=length(na.omit(hex.ext[[k]]))
    j=sum(na.omit(hex.ext[[k]]))
    if(j==0){
      hex.vals[k]=0
      next
    }
    if(n<11){
      if(j>0.7){hex.vals[k]=1}else{hex.vals[k]=0}
      next
    }
    if(n<41){
      if(j>0.5){hex.vals[k]=1}else{hex.vals[k]=0}
      next
    }
    if(n>40){
      if(j>0.3){hex.vals[k]=1}else{hex.vals[k]=0}
      next
    }
  }
  
  hex.data=as.data.frame(hex.vals)
  row.names(hex.data)=paste0("ID",1:length(hex))
  
  hex.data=SpatialPolygonsDataFrame(hex,hex.data)
  
  hex.write=st_as_sf(hex.data)
  
  st_write(hex.write,
           paste0(sdm.filepath,"hex/",name,".gpkg"))
  
  if(i==1){
    PAM=as.data.frame(hex.vals)
    colnames(PAM)[i]=name
  }else{
    PAM=cbind(PAM,hex.vals)
    index=which(colnames(PAM)=="hex.vals")
    colnames(PAM)[index]=name
  }
}

print('done')
```
```{r,eval=F}
print(errors)

row.names(PAM)=paste0('ID',1:nrow(PAM))
PAM2=t(PAM)

write.csv(PAM,paste0(sdm.path,"eco_PAM.csv"),row.names = T,quote=F)

print("done")
```
```
 [1] "Errors"                     "Amazona-ochrocephala"      
 [3] "Anthracothorax-nigricollis" "Ardea-cocoi"               
 [5] "Atalotriccus-pilaris"       "Cacicus-cela"              
 [7] "Campephilus-melanoleucos"   "Cantorchilus-leucotis"     
 [9] "Cercomacra-nigricans"       "Chaetura-brachyura"        
[11] "Chaetura-spinicaudus"       "Chalybura-buffonii"        
[13] "Cotinga-nattererii"         "Crotophaga-major"          
[15] "Dendroplex-picus"           "Euphonia-fulvicrissa"      
[17] "Glaucis-hirsutus"           "Icterus-chrysater"         
[19] "Juliamyia-julie"            "Lophornis-delattrei"       
[21] "Manacus-vitellinus"         "Momotus-subrufescens"      
[23] "Myiopagis-gaimardii"        "Myrmeciza-longipes"        
[25] "Notharchus-pectoralis"      "Oncostoma-olivaceum"       
[27] "Pachysylvia-aurantiifrons"  "Pitangus-lictor"           
[29] "Ramphocelus-dimidiatus"     "Rhynchocyclus-olivaceus"   
[31] "Sicalis-flaveola"           "Trogon-chionurus"          
[33] "Trogon-melanurus"          
[1] "done"
```

# 3 Summer models

# 3.1 Summer shapefiles

```{r,echo=F}
sdm.filepath=paste0("~/Dropbox/motmots/SDMs/SanJose_june/")
shp.filepath=paste0("~/Dropbox/motmots/shapefiles/")
```
```{r,eval=F}
# get list of files with points

gpkgs=list.files(shp.filepath,pattern=".gpkg")

y=readOGR(paste0(shp.filepath,gpkgs[2]))

crs.x=crs(y)
```

```{r,eval=F}
species=list.files(paste0(sdm.filepath,
                          "all.species"),
                   pattern="*.csv")
```
```{r,eval=F}
# preload shapefiles since we have only five

y1=readOGR(paste0(shp.filepath,gpkgs[1]))
y2=readOGR(paste0(shp.filepath,gpkgs[2]))
#y3=readOGR(paste0(shp.filepath,gpkgs[3]))
#y4=readOGR(paste0(shp.filepath,gpkgs[4]))
#y5=readOGR(paste0(shp.filepath,gpkgs[5]))

new.files="new.files"

for(i in 1:length(species)){
  x=read.csv(paste0(sdm.filepath,"all.species/",species[i]))
  
  nombre=strsplit(species[i],"[.]")[[1]][1]
  
  if(file.exists(paste0(shp.filepath,
                   "training_areas/",nombre,".gpkg"))==T){
      next
    }else{
      new.files=c(new.files,nombre)
    }
  
  x2=x%>%dplyr::select(LONGITUDE,LATITUDE)
  x.data2=SpatialPoints(x2,proj4string = crs.x)
  
  shp.list=NULL
  
  # for variable number of gpkgs files
  
  #for(k in 1:length(gpkgs)){
  #  y=readOGR(paste0(shp.filepath,gpkgs[k]))
  #  inside.m=sum(!is.na(over(x.data2,as(y,"SpatialPolygons"))))
  #  if(inside.m>0){shp.list[k]=1}else{shp.list[k]=0}
  #}
  
  # for this instance, since there are only five
  
  shp.list[1]=sum(!is.na(over(x.data2,as(y1,"SpatialPolygons"))))
  shp.list[2]=sum(!is.na(over(x.data2,as(y2,"SpatialPolygons"))))
  #shp.list[3]=sum(!is.na(over(x.data2,as(y3,"SpatialPolygons"))))
  #shp.list[4]=sum(!is.na(over(x.data2,as(y4,"SpatialPolygons"))))
  #shp.list[5]=sum(!is.na(over(x.data2,as(y5,"SpatialPolygons"))))
  
  shps=which(shp.list>0)
  
  if(length(shps)==0){next}
  
  if(length(shps)==1){
    # x.u.1=readOGR(paste0(shp.filepath,gpkgs[shps]))
    if(shps==1){assign('x.u.1',y1)}
    if(shps==2){assign('x.u.1',y2)}
    #if(shps==3){assign('x.u.1',y3)}
    #if(shps==4){assign('x.u.1',y4)}
    #if(shps==5){assign('x.u.1',y5)}
  }else{
    # merge shapefiles
    #x.u.1=readOGR(paste0(shp.filepath,gpkgs[shps[1]]))
    #for(k in 2:length(shps)){
    #  x.u.2=readOGR(paste0(shp.filepath,gpkgs[shps[k]]))
    #  x.u.1=gUnion(x.u.1,x.u.2)
    #}
    
    # memory saver below
    for(k in 1:length(shps)){
      if(k==1){
        if(shps[k]==1){assign('x.u.1',y1)}
        if(shps[k]==2){assign('x.u.1',y2)}
        #if(shps[k]==3){assign('x.u.1',y3)}
        #if(shps[k]==4){assign('x.u.1',y4)}
        #if(shps[k]==5){assign('x.u.1',y5)}
      }else{
        if(shps[k]==1){assign('x.u.2',y1)}
        if(shps[k]==2){assign('x.u.2',y2)}
        #if(shps[k]==3){assign('x.u.2',y3)}
        #if(shps[k]==4){assign('x.u.2',y4)}
        #if(shps[k]==5){assign('x.u.2',y5)}
        
        x.u.1=gUnion(x.u.1,x.u.2)
      }
    }
  }
  
  x.union=st_as_sf(x.u.1)
  
  st_write(x.union,
           paste0(shp.filepath,"training_areas/",nombre,".gpkg"),
           nombre)
}

print(new.files)
```
```
  [1] "new.files"                    "Amazona-ochrocephala"        
  [3] "Androdon-aequatorialis"       "Anthracothorax-nigricollis"  
  [5] "Anthus-lutescens"             "Aphanotriccus-audax"         
  [7] "Ara-chloropterus"             "Ara-severus"                 
  [9] "Ardea-cocoi"                  "Ardenna-grisea"              
 [11] "Arremon-atricapillus"         "Atalotriccus-pilaris"        
 [13] "Atlapetes-luteoviridis"       "Atticora-tibialis"           
 [15] "Basileuterus-ignotus"         "Brachygalba-salmoni"         
 [17] "Cacicus-cela"                 "Calidris-fuscicollis"        
 [19] "Campephilus-haematogaster"    "Campephilus-melanoleucos"    
 [21] "Campylorhynchus-albobrunneus" "Campylorhynchus-griseus"     
 [23] "Cantorchilus-leucopogon"      "Cantorchilus-leucotis"       
 [25] "Capito-maculicoronatus"       "Carpodectes-hopkei"          
 [27] "Caryothraustes-canadensis"    "Ceratopipra-erythrocephala"  
 [29] "Cercomacra-nigricans"         "Chaetura-brachyura"          
 [31] "Chaetura-spinicaudus"         "Chalybura-buffonii"          
 [33] "Chlorospingus-flavigularis"   "Chlorospingus-inornatus"     
 [35] "Chlorospingus-tacarcunae"     "Chlorothraupis-olivacea"     
 [37] "Chrysomus-icterocephalus"     "Circus-buffoni"              
 [39] "Cnipodectes-subbrunneus"      "Coccycua-minuta"             
 [41] "Colaptes-punctigula"          "Conirostrum-leucogenys"      
 [43] "Cotinga-nattererii"           "Cranioleuca-dissita"         
 [45] "Crotophaga-major"             "Cryptoleucopteryx-plumbea"   
 [47] "Cryptopipo-holochlora"        "Cyanerpes-caeruleus"         
 [49] "Dacnis-viguieri"              "Dendroplex-picus"            
 [51] "Donacobius-atricapilla"       "Euphonia-fulvicrissa"        
 [53] "Euphonia-xanthogaster"        "Fluvicola-pica"              
 [55] "Forpus-conspicillatus"        "Glaucis-hirsutus"            
 [57] "Goethalsia-bella"             "Goldmania-violiceps"         
 [59] "Gygis-alba"                   "Haplophaedia-aureliae"       
 [61] "Harpia-harpyja"               "Hemithraupis-flavicollis"    
 [63] "Herpsilochmus-rufimarginatus" "Heterospingus-xanthopygius"  
 [65] "Icterus-auricapillus"         "Icterus-chrysater"           
 [67] "Jacamerops-aureus"            "Juliamyia-julie"             
 [69] "Larosterna-inca"              "Larus-californicus"          
 [71] "Larus-delawarensis"           "Larus-dominicanus"           
 [73] "Larus-fuscus"                 "Leucophaeus-modestus"        
 [75] "Lophornis-delattrei"          "Machetornis-rixosa"          
 [77] "Manacus-vitellinus"           "Margarornis-bellulus"        
 [79] "Momotus-subrufescens"         "Morphnus-guianensis"         
 [81] "Myadestes-coloratus"          "Myiodynastes-chrysocephalus" 
 [83] "Myiopagis-caniceps"           "Myiopagis-gaimardii"         
 [85] "Myrmeciza-longipes"           "Myrmornis-torquata"          
 [87] "Myrmotherula-ignota"          "Myrmotherula-pacifica"       
 [89] "Nonnula-frontalis"            "Notharchus-pectoralis"       
 [91] "Nystalus-radiatus"            "Oncostoma-olivaceum"         
 [93] "Pachyramphus-homochrous"      "Pachyramphus-rufus"          
 [95] "Pachysylvia-aurantiifrons"    "Patagioenas-leucocephala"    
 [97] "Patagioenas-plumbea"          "Pelecanus-erythrorhynchos"   
 [99] "Phaethornis-anthophilus"      "Phaetusa-simplex"            
[101] "Philydor-fuscipenne"          "Phyllomyias-griseiceps"      
[103] "Phylloscartes-flavovirens"    "Piculus-callopterus"         
[105] "Piculus-chrysochloros"        "Pilherodius-pileatus"        
[107] "Pitangus-lictor"              "Poecilostreptus-palmeri"     
[109] "Polioptila-schistaceigula"    "Progne-elegans"              
[111] "Psarocolius-guatimozinus"     "Pseudobulweria-rostrata"     
[113] "Pyrrhura-picta"               "Quiscalus-lugubris"          
[115] "Ramphocelus-dimidiatus"       "Rhynchocyclus-olivaceus"     
[117] "Sapayoa-aenigma"              "Schiffornis-stenorhyncha"    
[119] "Selasphorus-ardens"           "Sicalis-flaveola"            
[121] "Sicalis-luteola"              "Sirystes-albogriseus"        
[123] "Sula-granti"                  "Tangara-fucosa"              
[125] "Tersina-viridis"              "Thamnophilus-nigriceps"      
[127] "Tolmomyias-flaviventris"      "Touit-dilectissimus"         
[129] "Trogon-chionurus"             "Trogon-melanurus"            
[131] "Vireolanius-eximius"          "Xenerpestes-minlosi"         
[133] "Zentrygon-goldmani"  
```

Some of the above are also just species that occur in Panama that are not in Costa Rica; therefore, not fully accurate.

## 2.2 Clipping Rasters

Next, we need to clip all rasters to the aforementioned training areas and ensure that we have the proper biogeographic envelopes applied for each species.

```{r,echo=F}
shp.path="~/Dropbox/motmots/shapefiles/training_areas/"
sdm.path="~/Dropbox/motmots/SDMs/SanJose_june/"
```

Note the SDM subdirectories are `output_all/75` for the whole files thresholded with 75% confidence and `clipped_SDM` for the clipped rasters.

```{r,eval=F}
sdm.list=list.files(paste0(sdm.path,"output/75"),pattern="*.tif")
shp.list=list.files(paste0(shp.path),pattern="*.gpkg")
```

We will not be clipping training areas for each species to correct for species richness patterns etc.

```{r,eval=F}
PAM=NULL

errors="Errors"

for(i in 1:length(sdm.list)){
  name=strsplit(sdm.list[i],"_")[[1]][1]

  if(length(which(shp.list%flike%name))<1){
    errors=c(errors,name)
    next
  }
  if(length(which(sdm.list%flike%name))<1){
    errors=c(errors,name)
    next
  }
  
  x=raster(paste0(sdm.path,"output/75/",sdm.list[i]))
  y=readOGR(paste0(shp.path,shp.list[which(shp.list%like%name)]))
  
  cr1=crop(x,y)
  cr2=raster::mask(cr1,y)
  crs(cr2)=crs(x.temp)
  cr3=projectRaster(cr2,x.temp,method='ngb',
                    res=(res(x)),crs=crs(x))
  
  # make everything zero so land cells become 0 when stacked
  cr3[is.na(cr3)]=0
  
  x.stack=stack(cr3,x.temp)
  cr4=calc(x.stack,fun=sum)
  
  # performed in previous iteration
  if(file.exists(paste0(sdm.path,'clipped_SDM/',
                              name,"_cropped.tif"))==F){
    writeRaster(cr4,
              filename=paste0(sdm.path,'clipped_SDM/',
                              name,"_cropped.tif"))
  }
  
  # project to hex
  # project to hex
  rm(x)
  rm(y)
  rm(cr1)
  rm(cr2)
  rm(cr3)
  rm(cr4)
  
  v.cr=velox(paste0(sdm.path,'clipped_SDM/',
                              name,"_cropped.tif"))
  
  hex.ext=v.cr$extract(hex)
  
  #hex.ext=raster::extract(cr4,hex)
  
  hex.vals=NULL
  
  for(k in 1:length(hex.ext)){
    n=length(na.omit(hex.ext[[k]]))
    j=sum(na.omit(hex.ext[[k]]))
    if(j==0){
      hex.vals[k]=0
      next
    }
    if(n<11){
      if(j>0.7){hex.vals[k]=1}else{hex.vals[k]=0}
      next
    }
    if(n<41){
      if(j>0.5){hex.vals[k]=1}else{hex.vals[k]=0}
      next
    }
    if(n>40){
      if(j>0.3){hex.vals[k]=1}else{hex.vals[k]=0}
      next
    }
  }
  
  hex.data=as.data.frame(hex.vals)
  row.names(hex.data)=paste0("ID",1:length(hex))
  
  hex.data=SpatialPolygonsDataFrame(hex,hex.data)
  
  hex.write=st_as_sf(hex.data)
  
  st_write(hex.write,
           paste0(sdm.filepath,"hex/",name,".gpkg"))
  
  #st_write(hex.data,
  #         paste0(shp.filepath,"training_areas/",nombre,".gpkg"),
  #         nombre)
  #writeOGR(hex.data,
  #         dsn=paste0(sdm.path,"hex/",name,"_hex-grid.gpkg"),
  #         layer=name,
  #         driver="SQLite")
  
  if(i==1){
    PAM=as.data.frame(hex.vals)
    colnames(PAM)[i]=name
  }else{
    PAM=cbind(PAM,hex.vals)
    index=which(colnames(PAM)=="hex.vals")
    colnames(PAM)[index]=name
  }
}

print("Done.")
```
```{r,eval=F}
print(errors)

row.names(PAM)=paste0('ID',1:nrow(PAM))
PAM2=t(PAM)

write.csv(PAM,paste0(sdm.path,"eco_PAM.csv"),row.names = T,quote=F)
```

```
 [1] "Errors"                    "Amazona-ochrocephala"     
 [3] "Atalotriccus-pilaris"      "Cacicus-cela"             
 [5] "Campephilus-melanoleucos"  "Cantorchilus-leucotis"    
 [7] "Chaetura-brachyura"        "Chaetura-spinicaudus"     
 [9] "Chalybura-buffonii"        "Crotophaga-major"         
[11] "Dendroplex-picus"          "Euphonia-fulvicrissa"     
[13] "Glaucis-hirsutus"          "Icterus-chrysater"        
[15] "Juliamyia-julie"           "Manacus-vitellinus"       
[17] "Momotus-subrufescens"      "Myiopagis-gaimardii"      
[19] "Myrmeciza-longipes"        "Notharchus-pectoralis"    
[21] "Oncostoma-olivaceum"       "Pachysylvia-aurantiifrons"
[23] "Pitangus-lictor"           "Ramphocelus-dimidiatus"   
[25] "Sicalis-flaveola"          "Trogon-chionurus"         
[27] "Trogon-melanurus"
```


# Full group unclipped - summer

```{r,eval=F}
sdm.path="~/Dropbox/motmots/SDMs/SanJose_june/"

sdm.list=list.files(paste0(sdm.path,"output/75"),pattern="*.tif")
```


```{r,eval=F}
PAM=NULL

errors="Errors"

for(i in 1:length(sdm.list)){
  name=strsplit(sdm.list[i],"_")[[1]][1]

  if(length(which(shp.list%flike%name))<1){
    errors=c(errors,name)
    next
  }
  if(length(which(sdm.list%flike%name))<1){
    errors=c(errors,name)
    next
  }
  
  #x=raster(paste0(sdm.path,"output_all/75/",sdm.list[i]))
  #y=readOGR(paste0(shp.path,shp.list[which(shp.list%like%name)]))
  
  #cr1=crop(x,y)
  #cr2=raster::mask(cr1,y)
  #crs(cr2)=crs(x.temp)
  #cr3=projectRaster(cr2,x.temp,method='ngb',
  #                  res=(res(x)),crs=crs(x))
  
  # make everything zero so land cells become 0 when stacked
  #cr3[is.na(cr3)]=0
  
  #x.stack=stack(cr3,x.temp)
  #cr4=calc(x.stack,fun=sum)
  
  #writeRaster(cr4,
  #            filename=paste0(sdm.path,'clipped_SDM/',
  #                            name,"_cropped.tif"))
  
  # project to hex
  # project to hex
  #rm(x)
  #rm(y)
  #rm(cr1)
  #rm(cr2)
  #rm(cr3)
  #rm(cr4)
  
  v.cr=velox(paste0(sdm.path,'output/75/',
                              sdm.list[i]))
  
  hex.ext=v.cr$extract(hex)
  
  #hex.ext=raster::extract(cr4,hex)
  
  hex.vals=NULL
  
  for(k in 1:length(hex.ext)){
    n=length(na.omit(hex.ext[[k]]))
    j=sum(na.omit(hex.ext[[k]]))
    if(j==0){
      hex.vals[k]=0
      next
    }
    if(n<11){
      if(j>0.7){hex.vals[k]=1}else{hex.vals[k]=0}
      next
    }
    if(n<41){
      if(j>0.5){hex.vals[k]=1}else{hex.vals[k]=0}
      next
    }
    if(n>40){
      if(j>0.3){hex.vals[k]=1}else{hex.vals[k]=0}
      next
    }
  }
  
  hex.data=as.data.frame(hex.vals)
  row.names(hex.data)=paste0("ID",1:length(hex))
  
  hex.data=SpatialPolygonsDataFrame(hex,hex.data)
  
  #hex.write=st_as_sf(hex.data)
  
  st_write(hex.write,
           paste0(sdm.filepath,"noclip/",name,".gpkg"))
  
  #st_write(hex.data,
  #         paste0(shp.filepath,"training_areas/",nombre,".gpkg"),
  #         nombre)
  #writeOGR(hex.data,
  #         dsn=paste0(sdm.path,"hex/",name,"_hex-grid.gpkg"),
  #         layer=name,
  #         driver="SQLite")
  
  if(i==1){
    PAM=as.data.frame(hex.vals)
    colnames(PAM)[i]=name
  }else{
    PAM=cbind(PAM,hex.vals)
    index=which(colnames(PAM)=="hex.vals")
    colnames(PAM)[index]=name
  }
}

print("Done")
```
```{r,eval=F}
print(errors)

row.names(PAM)=paste0('ID',1:nrow(PAM))
PAM2=t(PAM)

write.csv(PAM,paste0(sdm.path,"eco_null_PAM.csv"),row.names = T,quote=F)
```
```
 [1] "Errors"                    "Amazona-ochrocephala"     
 [3] "Atalotriccus-pilaris"      "Cacicus-cela"             
 [5] "Campephilus-melanoleucos"  "Cantorchilus-leucotis"    
 [7] "Chaetura-brachyura"        "Chaetura-spinicaudus"     
 [9] "Chalybura-buffonii"        "Crotophaga-major"         
[11] "Dendroplex-picus"          "Euphonia-fulvicrissa"     
[13] "Glaucis-hirsutus"          "Icterus-chrysater"        
[15] "Juliamyia-julie"           "Manacus-vitellinus"       
[17] "Momotus-subrufescens"      "Myiopagis-gaimardii"      
[19] "Myrmeciza-longipes"        "Notharchus-pectoralis"    
[21] "Oncostoma-olivaceum"       "Pachysylvia-aurantiifrons"
[23] "Pitangus-lictor"           "Ramphocelus-dimidiatus"   
[25] "Sicalis-flaveola"          "Trogon-chionurus"         
[27] "Trogon-melanurus"  
```


# Full group unclipped - winter

```{r,eval=F}
sdm.path="~/Dropbox/motmots/SDMs/MadreSelva_december/"

sdm.list=list.files(paste0(sdm.path,"output_all/75"),pattern="*.tif")
```


```{r,eval=F}
PAM=NULL

errors="Errors"

for(i in 411:length(sdm.list)){
  name=strsplit(sdm.list[i],"_")[[1]][1]

  if(length(which(shp.list%flike%name))<1){
    errors=c(errors,name)
    next
  }
  if(length(which(sdm.list%flike%name))<1){
    errors=c(errors,name)
    next
  }
  
  #x=raster(paste0(sdm.path,"output_all/75/",sdm.list[i]))
  #y=readOGR(paste0(shp.path,shp.list[which(shp.list%like%name)]))
  
  #cr1=crop(x,y)
  #cr2=raster::mask(cr1,y)
  #crs(cr2)=crs(x.temp)
  #cr3=projectRaster(cr2,x.temp,method='ngb',
  #                  res=(res(x)),crs=crs(x))
  
  # make everything zero so land cells become 0 when stacked
  #cr3[is.na(cr3)]=0
  
  #x.stack=stack(cr3,x.temp)
  #cr4=calc(x.stack,fun=sum)
  
  #writeRaster(cr4,
  #            filename=paste0(sdm.path,'clipped_SDM/',
  #                            name,"_cropped.tif"))
  
  # project to hex
  # project to hex
  #rm(x)
  #rm(y)
  #rm(cr1)
  #rm(cr2)
  #rm(cr3)
  #rm(cr4)
  
  v.cr=velox(paste0(sdm.path,'output_all/75/',
                              sdm.list[i]))
  
  hex.ext=v.cr$extract(hex)
  
  #hex.ext=raster::extract(cr4,hex)
  
  hex.vals=NULL
  
  for(k in 1:length(hex.ext)){
    n=length(na.omit(hex.ext[[k]]))
    j=sum(na.omit(hex.ext[[k]]))
    if(j==0){
      hex.vals[k]=0
      next
    }
    if(n<11){
      if(j>0.7){hex.vals[k]=1}else{hex.vals[k]=0}
      next
    }
    if(n<41){
      if(j>0.5){hex.vals[k]=1}else{hex.vals[k]=0}
      next
    }
    if(n>40){
      if(j>0.3){hex.vals[k]=1}else{hex.vals[k]=0}
      next
    }
  }
  
  hex.data=as.data.frame(hex.vals)
  row.names(hex.data)=paste0("ID",1:length(hex))
  
  hex.data=SpatialPolygonsDataFrame(hex,hex.data)
  
  hex.write=st_as_sf(hex.data)
  
  st_write(hex.write,
           paste0(sdm.path,"noclip/",name,".gpkg"))
  
  #st_write(hex.data,
  #         paste0(shp.filepath,"training_areas/",nombre,".gpkg"),
  #         nombre)
  #writeOGR(hex.data,
  #         dsn=paste0(sdm.path,"hex/",name,"_hex-grid.gpkg"),
  #         layer=name,
  #         driver="SQLite")
  
  if(i==1){
    PAM=as.data.frame(hex.vals)
    colnames(PAM)[i]=name
  }else{
    PAM=cbind(PAM,hex.vals)
    index=which(colnames(PAM)=="hex.vals")
    colnames(PAM)[index]=name
  }
}

print("Done")
```
```{r,eval=F}
print(errors)

row.names(PAM)=paste0('ID',1:nrow(PAM))
PAM2=t(PAM)

write.csv(PAM,paste0(sdm.path,"eco_null_PAM.csv"),row.names = T,quote=F)
```

```
 [1] "Errors"                     "Amazona-ochrocephala"      
 [3] "Anthracothorax-nigricollis" "Ardea-cocoi"               
 [5] "Atalotriccus-pilaris"       "Cacicus-cela"              
 [7] "Campephilus-melanoleucos"   "Cantorchilus-leucotis"     
 [9] "Cercomacra-nigricans"       "Chaetura-brachyura"        
[11] "Chaetura-spinicaudus"       "Chalybura-buffonii"        
[13] "Cotinga-nattererii"         "Crotophaga-major"          
[15] "Dendroplex-picus"           "Euphonia-fulvicrissa"      
[17] "Glaucis-hirsutus"           "Icterus-chrysater"         
[19] "Juliamyia-julie"            "Lophornis-delattrei"       
[21] "Manacus-vitellinus"         "Momotus-subrufescens"      
[23] "Myiopagis-gaimardii"        "Myrmeciza-longipes"        
[25] "Notharchus-pectoralis"      "Oncostoma-olivaceum"       
[27] "Pachysylvia-aurantiifrons"  "Pitangus-lictor"           
[29] "Ramphocelus-dimidiatus"     "Rhynchocyclus-olivaceus"   
[31] "Sicalis-flaveola"           "Trogon-chionurus"          
[33] "Trogon-melanurus"
```

